---
title: "Independent Project Introduction"
subtitle: "Phenotypic plasticity in *Eschscholzia californica* flowers"
author: "Shea Nolan"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

## Introduction

Flowering plants are vital to terrestrial ecosystems and human agriculture, making their responses to climate change a critical area of study. Flowers are essential for reproduction and influence population viability and fruit or seed yield. However, their physiological constraints remain understudied (Gleason, 2018). Research suggests flowers maintain their shape using a hydrostatic skeleton, which reduces carbon investment but relies on a positive water balance (Roddy et al., 2019). Despite requiring less carbon than leaves, flowers can lose significant water during drought due to petal surface conductance, potentially affecting floral display under stress (Roddy et al., 2023).

Phenotypic plasticity, the ability to express different phenotypes from a single genotype, helps plants adapt to environmental variability. This trait is critical in assessing plant resilience to climate change, particularly in species facing geographic or seasonal climate shifts (Matesanz et al., 2010; Valladares et al., 2014). *Eschscholzia californica*, the California golden poppy, has a long bloom period in a Mediterranean climate, spanning cool, wet winters and hot, dry summers. Herbarium data suggests *E. californica* flowering onset shifts with warming mean annual temperatures, and that drought induces maladaptive plasticity in reproductive fitness (Pearson et al., 2021; Ryan & Cleland, 2021). **Hypothesis:** I hypothesize that *E. californica* flowers exhibit plasticity in water-use traits to sustain floral displays across variable conditions.

```{r}
library(tidyverse)
library(rstatix)
library(car)

# load data
petal = read.csv("data/petal.csv")
petal$date = as.Date(petal$date, "%m/%d/%Y")
str(petal)
```

### Aim

Does **petal surface area** in *E. californica* plants vary with time?
I collected poppy petals from Mount Diablo State Park on a weekly basis from February 25th, 2025 to present. I scanned the petals and measured surface area in ImageJ. The following code chunk includes a table summarizing the mean and standard deviation for each of the petal samples by week. Note that there are smaller sample sizes (n) earlier in the season. This reflects both availability of the flowers and a shorter amount of time allocated for field collections.

```{r}
# table of petal stats for each week
petal_mean <- petal %>%
  group_by(date) %>%
  get_summary_stats(p_surfaceArea, type = "mean_sd")
petal_mean
```

Next, I called the ANOVA function in R with petal surface area and date as the inputs. I also ran this through an additional function that specifies a Type III ANOVA to account for differing sample sizes, but this doesn't seem to change the model output at all. The output suggests that at least one of the petal samples is different from the others in time.

```{r}
# type 3 ANOVA for unequal sample sizes
petal_aov <- aov(p_surfaceArea~factor(date),data=petal)
petal_aov3 <- Anova(petal_aov,type = 3)
petal_aov3
petal_aov
```

By estimating the R squared value, it seems that 36.2% of the variation in the data is explained by differences among treatment groups.

```{r}
petal_R2 = 567.83/(567.83 + 999.81)
petal_R2
```

I test the model assumptions here by plotting the residuals and running a Shapiro normality test. I used the residuals from the normal ANOVA test here as I was having trouble using the Type III ANOVA as the input. Visually, the data appears to follow a normal distribution, and the variance is approximately equal across groups. Each observation is independent from one another, new plants were selected haphazardly each week. A few outliers (69, 77, and 88) seem to weight the data a little unevenly. I also looked for equal variances with the Bartlett test, and it seems the petal data is well below the p-value threshold, signaling homogeneity in variences.

```{r}
# extracting model residuals
petal_aov_res1 <- residuals(petal_aov)

# converting aov_res into a dataframe for ggplot
petal_aov_res2 <- as.data.frame(petal_aov_res1)

# histogram of model residuals
ggplot(petal_aov_res2,aes(x=petal_aov_res1)) +
  geom_histogram(binwidth = 1)

# testing normality assumptions
plot(petal_aov)

# shapiro test
shapiro.test(petal_aov_res1)

# bartlett test
bartlett.test(p_surfaceArea~factor(date), data=petal) 
```
In the following code chunk I call a Tukey test function to determine which petals samples signal as distinct from others in time. I also used the normal one-way ANOVA for this, as I did not have the time to try and make the function parse the Type III ANOVA. I am also not sure how to adjust parameters in the plot function, so unfortunately the y axis is not readable. Looking at the table, it seems the most distinct samples are from April 29th and April 15th.

```{r}
# Tukey test to see which dates are significantly different
t <- TukeyHSD(petal_aov)
t

# Visualization of the Tukey test
plot(t)
```

A boxplot follows for the petal data I have collected thus far. It seems there may be a trend in decreasing petal size over time. I plan to collect more petals, and I'd like to try and follow more randomized methods of petal collection rather than rely on a haphazard approach.

```{r}
my_title <- expression(paste("Variation in ", italic("Eschscholzia californica"), " petal surface area over time"))

# boxplot of petal samples by week
ggplot(petal, aes(group = date, x = date, y = p_surfaceArea)) +
  geom_boxplot() +
  labs(title = my_title, x = "Time of the year", y = "Petal surface area (cm2)")
```

## Literature cited

Gleason, S. M. (2018). A blooming interest in the hydraulic traits of flowers. *Plant, Cell & Environment, 41*(10), 2247–2249. <https://doi.org/10.1111/pce.13345>

Matesanz, S., Gianoli, E., & Valladares, F. (2010). Global change and the evolution of phenotypic plasticity in plants. *Annals of the New York Academy of Sciences, 1206*(1), 35–55. <https://doi.org/10.1111/j.1749-6632.2010.05704.x>

Pearson, K. D., Love, N. L. R., Ramirez-Parada, T., Mazer, S. J., & Yost, J. M. (2021). Phenological trends in the California poppy (*Eschscholzia californica*): digitized herbarium specimens reveal intraspecific variation in the sensitivity of flowering date to climate change. *Madroño, 68*(4). <https://doi.org/10.3120/0024-9637-68.4.343>

Roddy, A. B., Guilliams, C. M., Fine, P. V. A., Mambelli, S., Dawson, T. E., & Simonin, K. A. (2023). Flowers are leakier than leaves but cheaper to build. *New Phytologist, 239*(6), 2076–2082. <https://doi.org/10.1111/nph.19104>

Roddy, A. B., Jiang, G., Cao, K., Simonin, K. A., & Brodersen, C. R. (2019). Hydraulic traits are more diverse in flowers than in leaves. *New Phytologist, 223*(1), 193–203. <https://doi.org/10.1111/nph.15749>

Ryan, E. M., & Cleland, E. E. (2021). Clinal variation in phenological traits and fitness responses to drought across the native range of California poppy. *Climate Change Ecology, 2*, 100021. <https://doi.org/10.1016/j.ecochg.2021.100021>

Valladares, F., Matesanz, S., Guilhaumon, F., Araújo, M. B., Balaguer, L., Benito-Garzón, M., Cornwell, W., Gianoli, E., van Kleunen, M., Naya, D. E., Nicotra, A. B., Poorter, H., & Zavala, M. A. (2014). The effects of phenotypic plasticity and local adaptation on forecasts of species range shifts under climate change. *Ecology Letters, 17*(11), 1351–1364. <https://doi.org/10.1111/ele.12348>
